---
title: "csbio100 project"
output: html_document
date: "2026-02-20"
---
PCA → genetic differentiation
CNV heatmap → functional pharmacogenomic variation

# Data Wrangling: SNP data from 1000 genomes dataset
```{r}
library(bigsnpr)
library(bigstatsr)

# downloads these into data_1000G/
# 1000G_phase3_common_norel.bed/.bim/.fam/.fam2
bigsnpr::download_1000G("data_1000G")

bedfile <- "data_1000G/1000G_phase3_common_norel.bed"

# IMPORTANT:
#rds <-  snp_readBed(bedfile)
#obj <- snp_attach(rds)
rds <- sub("\\.bed$", ".rds", bedfile)
obj <- snp_attach(rds)

G   <- obj$genotypes
map <- obj$map
fam <- obj$fam

# read population labels
fam2 <- read.table("data_1000G/1000G_phase3_common_norel.fam2",
                   header = TRUE, sep = "\t", stringsAsFactors = FALSE)

# quick sanity checks
stopifnot(all(c("sample.ID","Population","Super.Population") %in% names(fam2)))

```

1. PCA 
# 1a. PCA on Superpopulations
--> identify most separated superpopulations
```{r}
start_bp <- 41500000
end_bp   <- 43500000

# SNP SUBSET (locus window) 
ind.locus <- which(map$chromosome == 22 &
                     map$physical.pos >= start_bp &
                     map$physical.pos <= end_bp)

# MAF filter
maf <- snp_MAF(G, ind.col = ind.locus)
ind.locus2 <- ind.locus[maf >= 0.05 & maf <= 0.95]

cat("SNPs after MAF:", length(ind.locus2), "\n")

# PCA ON INDIVIDUALS
X <- as.matrix(G[, ind.locus2])

# drop zero-variance SNPs
keep <- apply(X, 2, var) > 0
X <- X[, keep, drop = FALSE]
X <- scale(X)

# PCA
pca2 <- prcomp(X, center = FALSE, scale. = FALSE)

# MERGE LABELS 
pca2_df <- data.frame(
  sample.ID = fam$sample.ID,
  PC1 = pca2$x[,1],
  PC2 = pca2$x[,2],
  stringsAsFactors = FALSE
)

pca2_df <- merge(pca2_df, fam2, by = "sample.ID", all.x = TRUE)

#  PLOT (colored by superpop) 
sp <- factor(pca2_df$Super.Population)

plot(pca2_df$PC1, pca2_df$PC2,
     col = sp, pch = 19,
     xlab = "PC1", ylab = "PC2",
     main = paste0("chr22:", start_bp, "-", end_bp, " (CYP2D6 locus region)"))

legend("topright", legend = levels(sp),
       col = seq_along(levels(sp)), pch = 19, cex = 0.8)
```

# 1b. PCA on Subpopulations (AFR and EAS) 
--> identify most separated subpopulations
```{r}
# SUBSET TO AFR + EAS INDIVIDUALS
pops_keep <- c("AFR", "EAS")
sub_ids <- fam2$sample.ID[fam2$Super.Population %in% pops_keep]
ind_rows <- which(fam$sample.ID %in% sub_ids)

# PCA RECOMPUTED ON ONLY AFR+EAS INDIVIDUALS 
# reuse the SAME SNP set ind.locus2 from Script 2 (important)
X_sub <- as.matrix(G[ind_rows, ind.locus2])

keep <- apply(X_sub, 2, var) > 0
X_sub <- X_sub[, keep, drop = FALSE]
X_sub <- scale(X_sub)

pca_sub <- prcomp(X_sub, center = FALSE, scale. = FALSE)

# dataframe for individuals in AFR+EAS
pca_sub_df <- data.frame(
  sample.ID = fam$sample.ID[ind_rows],
  PC1 = pca_sub$x[,1],
  PC2 = pca_sub$x[,2],
  stringsAsFactors = FALSE
)

pca_sub_df <- merge(pca_sub_df, fam2, by = "sample.ID")

# POPULATION CENTROIDS (GBR/FIN/etc — but here only AFR+EAS pops exist)
centroids_pop <- aggregate(cbind(PC1, PC2) ~ Population,
                           data = pca_sub_df,
                           mean)

# add superpopulation label for each Population
centroids_pop <- merge(centroids_pop,
                       unique(fam2[, c("Population", "Super.Population")]),
                       by = "Population")

# PLOT CENTROIDS COLORED BY SUPERPOP
cols <- ifelse(centroids_pop$Super.Population == "AFR", "black", "red")

plot(centroids_pop$PC1, centroids_pop$PC2,
     col = cols, pch = 19, cex = 2,
     xlab = "PC1", ylab = "PC2",
     main = "Population-Level PCA (AFR vs EAS)")

text(centroids_pop$PC1, centroids_pop$PC2,
     labels = centroids_pop$Population,
     pos = 3, cex = 0.9)

legend("topright", legend = c("AFR", "EAS"),
       col = c("black", "red"), pch = 19)

# MOST SEPARATED AFR vs EAS POPULATION PAIR (centroid distance) 
afr_cent <- subset(centroids_pop, Super.Population == "AFR")
eas_cent <- subset(centroids_pop, Super.Population == "EAS")

dist_matrix <- matrix(NA, nrow = nrow(afr_cent), ncol = nrow(eas_cent))
rownames(dist_matrix) <- afr_cent$Population
colnames(dist_matrix) <- eas_cent$Population

for (i in 1:nrow(afr_cent)) {
  for (j in 1:nrow(eas_cent)) {
    dist_matrix[i, j] <- sqrt(
      (afr_cent$PC1[i] - eas_cent$PC1[j])^2 +
        (afr_cent$PC2[i] - eas_cent$PC2[j])^2
    )
  }
}

max_idx <- which(dist_matrix == max(dist_matrix), arr.ind = TRUE)
afr_pop <- rownames(dist_matrix)[max_idx[1]]
eas_pop <- colnames(dist_matrix)[max_idx[2]]

cat("Most separated AFR–EAS pair:", afr_pop, "vs", eas_pop, "\n")
cat("Distance =", max(dist_matrix), "\n")

```



2. Allele Frequency Heatmaps
```{r}
# CYP2D6 region (slightly padded)
start_bp <- 42100000
end_bp   <- 42250000

ind.cyp <- which(map$chromosome == 22 &
                 map$physical.pos >= start_bp &
                 map$physical.pos <= end_bp)
length(ind.cyp)

maf <- bigsnpr::snp_MAF(G, ind.col = ind.cyp)
# keep informative SNPs
ind.cyp2 <- ind.cyp[maf >= 0.05 & maf <= 0.95]
cat("SNPs retained:", length(ind.cyp2), "\n")
```

```{r}
# extract genotype submatrix
X <- as.matrix(G[, ind.cyp2])

# add population labels
pop_labels <- fam2$Super.Population[
  match(fam$sample.ID, fam2$sample.ID)
]
unique(pop_labels)
```

# 2a. Allele Frequency Heatmap (Superpopulations)
```{r}
# compute allele frequency per superpopulation
pop_levels <- unique(pop_labels)

af_matrix <- matrix(NA,
                    nrow = length(pop_levels),
                    ncol = ncol(X))

rownames(af_matrix) <- pop_levels
colnames(af_matrix) <- paste0("SNP_", 1:ncol(X))

for (p in pop_levels) {
  inds <- which(pop_labels == p)
  af_matrix[p, ] <- colMeans(X[inds, ], na.rm = TRUE) / 2
}
```
Heatmap (all Superpopulations)
```{r}
#install.packages("pheatmap")
library(pheatmap)

pheatmap(af_matrix,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         color = colorRampPalette(c("navy", "white", "firebrick"))(100),
         main = "CYP2D6 Allele Frequency Heatmap\nSuperpopulations")
```
Heatmap (AFR & EAS only)
```{r}
library(pheatmap)

# Subset matrix to AFR and EAS
af_matrix_subset <- af_matrix[c("AFR", "EAS"), ]

# Plot heatmap
pheatmap(af_matrix_subset,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         color = colorRampPalette(c("navy", "white", "firebrick"))(100),
         main = "CYP2D6 Allele Frequency Heatmap\nAFR vs EAS")
```

# 2b. Allele Frequency Heatmap (Subpopulations of AFR & EAS)
```{r}
# Subpopulation labels 
pop_sub <- fam2$Population[
  match(fam$sample.ID, fam2$sample.ID)
]

super_sub <- fam2$Super.Population[
  match(fam$sample.ID, fam2$sample.ID)
]

# keep AFR + EAS individuals
keep_inds <- which(super_sub %in% c("AFR", "EAS"))

X_sub <- as.matrix(G[keep_inds, ind.cyp2])

pop_sub_keep <- pop_sub[keep_inds]
```

```{r}
subpops <- unique(pop_sub_keep)

af_matrix <- matrix(NA,
                    nrow = length(subpops),
                    ncol = ncol(X_sub))

rownames(af_matrix) <- subpops
colnames(af_matrix) <- paste0("SNP_", 1:ncol(X_sub))

for (p in subpops) {
  inds <- which(pop_sub_keep == p)
  af_matrix[p, ] <- colMeans(X_sub[inds, ], na.rm = TRUE) / 2
}

var_keep <- apply(af_matrix, 2, var) > 0.001
af_matrix <- af_matrix[, var_keep]
```
Heatmap
```{r}
library(pheatmap)
pheatmap(af_matrix,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         #show_colnames = FALSE, #SNPs
         color = colorRampPalette(c("navy", "white", "firebrick"))(100),
         main = "CYP2D6 Allele Frequency\nAFR vs EAS Subpopulations")
```




3. Bootstrap Resampling for Cluster Stability (ARI)
Each bootstrap: Resample individuals → Recompute PCA → Recluster → Compare with truth
Then we compute: CI95% = [Q25%, Q75%]
```{r}
library(mclust)
set.seed(123)

B <- 1000   # use 10000 for final submission
n <- length(ind_rows)

# ORIGINAL PCA + CLUSTERING
X_orig <- as.matrix(G[ind_rows, ind.locus2])
keep <- apply(X_orig, 2, var) > 0
X_orig <- X_orig[, keep, drop = FALSE]
X_orig <- scale(X_orig)

pca_orig <- prcomp(X_orig, center = FALSE, scale. = FALSE)

# Use first 2 PCs for clustering
pc_orig <- pca_orig$x[, 1:2]

# k-means clustering (k=2)
km_orig <- kmeans(pc_orig, centers = 2, nstart = 20)
orig_clusters <- km_orig$cluster

# True labels (AFR vs EAS)
true_labels <- fam2$Super.Population[match(fam$sample.ID[ind_rows], fam2$sample.ID)]

# Convert AFR/EAS to numeric labels
true_labels_num <- as.numeric(factor(true_labels))

# ARI between original clustering and true labels
ari_original <- adjustedRandIndex(orig_clusters, true_labels_num)
cat("Original ARI:", ari_original, "\n")
```
Bootstrap
```{r}
ari_vals <- numeric(B)
for (b in 1:B) {
  # resample individuals with replacement
  boot_idx <- sample(1:n, n, replace = TRUE)
  
  X_boot <- X_orig[boot_idx, ]
  
  # recompute PCA
  pca_boot <- prcomp(X_boot, center = FALSE, scale. = FALSE)
  pc_boot <- pca_boot$x[, 1:2]
  
  # recluster
  km_boot <- kmeans(pc_boot, centers = 2, nstart = 10)
  boot_clusters <- km_boot$cluster
  
  # true labels for bootstrap sample
  boot_true <- true_labels_num[boot_idx]
  
  # compute ARI
  ari_vals[b] <- adjustedRandIndex(boot_clusters, boot_true)
}
```
Summary Statistics
```{r}
mean_ari <- mean(ari_vals)
ci <- quantile(ari_vals, c(0.025, 0.975))

cat("Bootstrap Mean ARI:", mean_ari, "\n")
cat("95% CI:", ci[1], "-", ci[2], "\n")

# Plot Distribution
hist(ari_vals, breaks = 40,
     main = "Bootstrap Distribution of ARI (AFR vs EAS)",
     xlab = "Adjusted Rand Index",
     col = "lightblue", border = "white")

abline(v = ci, col = "red", lwd = 2)
abline(v = mean_ari, col = "darkblue", lwd = 2)
```




BONUS: UMAP
```{r}
library(uwot)

set.seed(123)

# Use same scaled genotype matrix from PCA
umap_res <- umap(
  X,
  n_neighbors = 15,
  min_dist = 0.1,
  metric = "euclidean"
)

umap_df <- data.frame(
  sample.ID = fam$sample.ID,
  UMAP1 = umap_res[,1],
  UMAP2 = umap_res[,2],
  stringsAsFactors = FALSE
)

umap_df <- merge(umap_df, fam2, by = "sample.ID")
```

```{r}
sp <- factor(umap_df$Super.Population)

plot(umap_df$UMAP1, umap_df$UMAP2,
     col = sp,
     pch = 19,
     xlab = "UMAP1",
     ylab = "UMAP2",
     main = "UMAP of CYP2D6 Region (1000 Genomes Superpopulations)")

legend("topright",
       legend = levels(sp),
       col = seq_along(levels(sp)),
       pch = 19)
```




4. CNV Analysis
--> see if CYP2D6 functional difference matches the genetic separation.
```{r}
library(vcfR)
vcf <- read.vcfR("1000G_SV/CYP2D6_SV.vcf.gz")
head(vcf@fix)
```

```{r}
var_rows <- which(apply(gt, 1, function(x) length(unique(x)) > 1))
gt_var <- gt[var_rows, ]

cnv_numeric <- apply(gt_var, 2, function(x) {
  mean(ifelse(x == "0|0", 2,
       ifelse(x %in% c("0|1","1|0"), 3,
       ifelse(x == "1|1", 4, NA))), na.rm = TRUE)
})

cnv_df <- data.frame(
  sample.ID = colnames(gt),
  copy_number = cnv_numeric
)

cnv_df <- merge(cnv_df, fam2, by = "sample.ID")

cnv_focus <- cnv_df %>%
  dplyr::filter(Population %in% c("ESN", "CDX"))
```

```{r}
cnv_summary <- cnv_df %>%
  filter(Super.Population %in% c("AFR","EAS")) %>%
  group_by(Population, Super.Population) %>%
  summarise(
    mean_CN = mean(copy_number),
    sd_CN = sd(copy_number),
    .groups="drop"
  )

ggplot(cnv_summary,
       aes(x = Population, y = mean_CN, color = Super.Population)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = mean_CN - sd_CN,
                    ymax = mean_CN + sd_CN),
                width = 0.2) +
  theme_minimal() +
  labs(
    title = "Mean CYP2D6 Copy Number by Population",
    y = "Mean Copy Number"
  )
```





BONUS: Causual Inference
1. Instrument (IV): CYP2D6 genotype (0/1/2 alleles).
2. Exposure: CYP2D6 activity (or CYP2D6 genotype coded numerically).
3. Outcome: PK measurement (DV, drug concentration at time points).
Run Mendelian Randomization (MR) to estimate the causal effect of CYP2D6 variation on PK outcomes.
```{r}
library(dplyr)
library(ggplot2)

# Load PK datasets 
nigerian <- read.csv("synthetic_nigerian_popPK.csv")
chinese  <- read.csv("synthetic_chinese_popPK.csv")

nigerian$POP <- "Nigerian"
chinese$POP  <- "Chinese"

pk_data <- bind_rows(nigerian, chinese)

# Compute PK summary per individual (AUC via trapezoidal rule)
pk_summary <- pk_data %>%
  group_by(ID, POP) %>%
  arrange(TIME) %>%
  summarise(
    AUC = sum(diff(TIME) * (head(DV,-1) + tail(DV,-1)) / 2, na.rm = TRUE),
    .groups = "drop"
  )

# Assign CYP2D6 activity levels 
# 0 = poor metabolizer
# 1 = intermediate
# 2 = normal / ultrarapid

set.seed(123)

pk_summary$CYP2D6_activity <- sample(
  0:2,
  size = nrow(pk_summary),
  replace = TRUE
)

# Check distribution
table(pk_summary$CYP2D6_activity)

# Mendelian Randomization Model
# CYP2D6 activity (instrument) → PK outcome (AUC)
# Adjust for population
mr_model <- lm(AUC ~ CYP2D6_activity + POP, data = pk_summary)

summary(mr_model)

# 5. Extract causal estimate
beta <- coef(mr_model)["CYP2D6_activity"]

se <- summary(mr_model)$coefficients["CYP2D6_activity","Std. Error"]

ci_lower <- beta - 1.96 * se
ci_upper <- beta + 1.96 * se

cat("Causal effect estimate (CYP2D6 activity -> AUC):", beta, "\n")
cat("95% CI:", ci_lower, "-", ci_upper, "\n")

# Create MR results dataframe
mr_results <- data.frame(
  Instrument = "CYP2D6 activity",
  Estimate = beta,
  SE = se,
  CI_low = ci_lower,
  CI_high = ci_upper
)

# 7. Plot causal effect estimate
ggplot(mr_results, aes(x = Instrument, y = Estimate)) +
  geom_point(size = 4) +
  geom_errorbar(
    aes(ymin = CI_low, ymax = CI_high),
    width = 0.2
  ) +
  theme_minimal() +
  ylab("Causal Effect on AUC") +
  ggtitle("Mendelian Randomization: CYP2D6 Activity → Alcohol Exposure (AUC)")
```



