---
title: "heatmap"
output: html_document
date: "2026-02-20"
---

allele frequency heatmap
```{r}
# CYP2D6 region (slightly padded)
start_bp <- 42100000
end_bp   <- 42250000

ind.cyp <- which(map$chromosome == 22 &
                 map$physical.pos >= start_bp &
                 map$physical.pos <= end_bp)

length(ind.cyp)
```

```{r}
maf <- bigsnpr::snp_MAF(G, ind.col = ind.cyp)

# keep informative SNPs
ind.cyp2 <- ind.cyp[maf >= 0.05 & maf <= 0.95]

cat("SNPs retained:", length(ind.cyp2), "\n")
```


```{r}
# extract genotype submatrix
X <- as.matrix(G[, ind.cyp2])

# add population labels
pop_labels <- fam2$Super.Population[
  match(fam$sample.ID, fam2$sample.ID)
]

unique(pop_labels)
```

```{r}
# compute allele frequency per superpopulation
pop_levels <- unique(pop_labels)

af_matrix <- matrix(NA,
                    nrow = length(pop_levels),
                    ncol = ncol(X))

rownames(af_matrix) <- pop_levels
colnames(af_matrix) <- paste0("SNP_", 1:ncol(X))

for (p in pop_levels) {
  inds <- which(pop_labels == p)
  af_matrix[p, ] <- colMeans(X[inds, ], na.rm = TRUE) / 2
}
```

```{r}
#install.packages("pheatmap")
library(pheatmap)

pheatmap(af_matrix,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         color = colorRampPalette(c("navy", "white", "firebrick"))(100),
         main = "CYP2D6 Allele Frequency Heatmap\nSuperpopulations")
```


```{r}
library(pheatmap)

# Subset matrix to AFR and EAS
af_matrix_subset <- af_matrix[c("AFR", "EAS"), ]

# Plot heatmap
pheatmap(af_matrix_subset,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         color = colorRampPalette(c("navy", "white", "firebrick"))(100),
         main = "CYP2D6 Allele Frequency Heatmap\nAFR vs EAS")
```



Average frequency of non-reference (variant) alleles across CYP2D6 SNPs within each superpopulation.
--> How genetically different populations are at the CYP2D6 locus.
--> Quantifies population-level genetic variation (beyond PCA visualization).
```{r}
# use your previously filtered ind.cyp2
X <- as.matrix(G[, ind.cyp2])

pop_labels <- fam2$Super.Population[
  match(fam$sample.ID, fam2$sample.ID)
]

pop_levels <- unique(pop_labels)
mean_burden <- data.frame()

for (p in pop_levels) {
  inds <- which(pop_labels == p)
  mean_alt <- mean(colMeans(X[inds, ], na.rm = TRUE) / 2)
  mean_burden <- rbind(mean_burden,
                       data.frame(Population = p,
                                  MeanAlternateAllele = mean_alt))
}

library(ggplot2)
ggplot(mean_burden,
       aes(x = Population,
           y = MeanAlternateAllele,
           fill = Population)) +
  geom_bar(stat = "identity") +
  labs(title = "Mean CYP2D6 Alternate Allele Burden by Superpopulation",
       y = "Mean Alternate Allele Frequency") +
  theme_minimal()
```


```{r}
# Subpopulation labels 
pop_sub <- fam2$Population[
  match(fam$sample.ID, fam2$sample.ID)
]

super_sub <- fam2$Super.Population[
  match(fam$sample.ID, fam2$sample.ID)
]

# keep AFR + EAS individuals
keep_inds <- which(super_sub %in% c("AFR", "EAS"))

X_sub <- as.matrix(G[keep_inds, ind.cyp2])

pop_sub_keep <- pop_sub[keep_inds]
```

```{r}
subpops <- unique(pop_sub_keep)

af_matrix <- matrix(NA,
                    nrow = length(subpops),
                    ncol = ncol(X_sub))

rownames(af_matrix) <- subpops
colnames(af_matrix) <- paste0("SNP_", 1:ncol(X_sub))

for (p in subpops) {
  inds <- which(pop_sub_keep == p)
  af_matrix[p, ] <- colMeans(X_sub[inds, ], na.rm = TRUE) / 2
}

var_keep <- apply(af_matrix, 2, var) > 0.001
af_matrix <- af_matrix[, var_keep]
```

Alternate allele frequency at each SNP.
```{r}
library(pheatmap)
pheatmap(af_matrix,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         #show_colnames = FALSE, #SNPs
         color = colorRampPalette(c("navy", "white", "firebrick"))(100),
         main = "CYP2D6 Allele Frequency\nAFR vs EAS Subpopulations")
```


```{r}
# keep only ESN and CDX
af_matrix_pair <- af_matrix[c("ESN", "CDX"), ]

# check dimensions
dim(af_matrix_pair)

# difference in allele frequency between the two populations
snp_diff <- abs(af_matrix_pair["ESN", ] - af_matrix_pair["CDX", ])

# select top 30 SNPs
top_snps <- order(snp_diff, decreasing = TRUE)#[1:20]
af_matrix_pair_top <- af_matrix_pair[, top_snps]

pheatmap(af_matrix_pair_top,
         cluster_rows = FALSE,  # only two rows, no need to cluster
         cluster_cols = TRUE,
         color = colorRampPalette(c("navy", "white", "firebrick"))(100),
         main = "CYP2D6 Allele Frequency\nESN vs CDX Subpopulations")
         #\n(Top 30 SNPs by Allele Frequence Difference)")
    

```

```{r}
af_matrix_pair_all <- af_matrix[c("ESN", "CDX"), ]
pheatmap(af_matrix_pair_all,
         cluster_rows = FALSE,  # only 2 populations
         cluster_cols = TRUE,   # cluster SNPs to show pattern
         show_colnames = FALSE, # hide individual SNP labels to avoid blob
         color = colorRampPalette(c("navy", "white", "firebrick"))(100),
         main = "CYP2D6 Allele Frequency\nESN vs CDX Subpopulations")
```


Top 20 SNPs by Allele Frequence Difference: 
Subpopulations 
```{r}
# Step 1: columns of af_matrix_pair correspond to which columns of af_matrix
# af_matrix_pair <- af_matrix[c("ESN","CDX"), var_keep]
var_keep_cols <- which(var_keep)  # positions in ind.cyp2

# Step 2: top 20 SNPs by ESN vs CDX diff (relative to af_matrix_pair)
top20_sub <- order(snp_diff_sub, decreasing = TRUE)[1:20]

# Step 3: map back to ind.cyp2
top20_ind_cyp2 <- ind.cyp2[var_keep_cols[top20_sub]]

# Step 4: map back to map for SNP_ID and Position
snp_ids_top20 <- map$marker.ID[top20_ind_cyp2]
snp_positions_top20 <- map$physical.pos[top20_ind_cyp2]

# Step 5: Build table
top20_snps_table <- data.frame(
  SNP_index = top20_sub,
  SNP_ID = snp_ids_top20,
  Position = snp_positions_top20,
  ESN_vs_CDX_diff = snp_diff_sub[top20_sub]
)

top20_snps_table
```



Overlap: Super vs. Sub populations
```{r}
# AFR vs EAS
keep_inds_super <- which(super_sub %in% c("AFR", "EAS"))
X_super <- as.matrix(G[keep_inds_super, ind.cyp2])
pop_super_keep <- super_sub[keep_inds_super]

superpops <- unique(pop_super_keep)

af_matrix_super <- sapply(superpops, function(p) {
  inds <- which(pop_super_keep == p)
  colMeans(X_super[inds, ], na.rm = TRUE)/2
}) 
af_matrix_super <- t(af_matrix_super)
rownames(af_matrix_super) <- superpops
colnames(af_matrix_super) <- paste0("SNP_", 1:ncol(af_matrix_super))

# SNP differences
snp_diff_super <- abs(af_matrix_super["AFR", ] - af_matrix_super["EAS", ])
top_snps_super <- order(snp_diff_super, decreasing = TRUE)[1:20]

# ESN vs CDX already computed as af_matrix_pair
snp_diff_sub <- abs(af_matrix_pair["ESN", ] - af_matrix_pair["CDX", ])
top_snps_sub <- order(snp_diff_sub, decreasing = TRUE)[1:20]

# Overlap of top SNPs
overlap_snps <- intersect(top_snps_super, top_snps_sub)
cat("Number of SNPs overlapping top 20 in both comparisons:", length(overlap_snps), "\n")
overlap_snps

# Optional: get physical positions of these SNPs
snp_positions <- map$physical.pos[ind.cyp2]
data.frame(
  SNP_index = overlap_snps,
  Position = snp_positions[overlap_snps],
  AFR_vs_EAS_diff = snp_diff_super[overlap_snps],
  ESN_vs_CDX_diff = snp_diff_sub[overlap_snps]
)
```

```{r}
# Overlapping SNP indices from your analysis
overlap_inds <- c(27, 35, 38, 32, 44, 21, 6, 46, 45)

# SNP IDs and positions
snp_ids <- map$marker.ID[ind.cyp2][overlap_inds]
snp_positions <- map$physical.pos[ind.cyp2][overlap_inds]

# Build table
overlap_snps_table <- data.frame(
  SNP_index = overlap_inds,
  SNP_ID = snp_ids,
  Position = snp_positions,
  AFR_vs_EAS_diff = snp_diff_super[overlap_inds],
  ESN_vs_CDX_diff = snp_diff_sub[overlap_inds]
)

overlap_snps_table
```






