---
title: "csbio100 project"
output: html_document
date: "2026-02-20"
---
PCA → genetic differentiation
CNV heatmap → functional pharmacogenomic variation

# Data Wrangling: SNP data from 1000 genomes dataset
```{r}
library(bigsnpr)
library(bigstatsr)

# downloads these into data_1000G/
# 1000G_phase3_common_norel.bed/.bim/.fam/.fam2
bigsnpr::download_1000G("data_1000G")

bedfile <- "data_1000G/1000G_phase3_common_norel.bed"

# IMPORTANT:
#rds <-  snp_readBed(bedfile)
#obj <- snp_attach(rds)
rds <- sub("\\.bed$", ".rds", bedfile)
obj <- snp_attach(rds)

G   <- obj$genotypes
map <- obj$map
fam <- obj$fam

# read population labels
fam2 <- read.table("data_1000G/1000G_phase3_common_norel.fam2",
                   header = TRUE, sep = "\t", stringsAsFactors = FALSE)

# quick sanity checks
stopifnot(all(c("sample.ID","Population","Super.Population") %in% names(fam2)))

```

1. PCA 
# 1a. PCA on Superpopulations
--> identify most separated superpopulations
CYP2D6 location (GRCh38): chr22:42126499–42130947
chr22:42126499–42130947
```{r}
start_bp <- 41600000
end_bp   <- 43300000

# SNP SUBSET (locus window) 
ind.locus <- which(map$chromosome == 22 &
                     map$physical.pos >= start_bp &
                     map$physical.pos <= end_bp)

# MAF filter
maf <- snp_MAF(G, ind.col = ind.locus)
ind.locus2 <- ind.locus[maf >= 0.05 & maf <= 0.95]

cat("SNPs after MAF:", length(ind.locus2), "\n")

# PCA ON INDIVIDUALS
X <- as.matrix(G[, ind.locus2])

# drop zero-variance SNPs
keep <- apply(X, 2, var) > 0
X <- X[, keep, drop = FALSE]
X <- scale(X)

# PCA
pca2 <- prcomp(X, center = FALSE, scale. = FALSE)

# MERGE LABELS 
pca2_df <- data.frame(
  sample.ID = fam$sample.ID,
  PC1 = pca2$x[,1],
  PC2 = pca2$x[,2],
  stringsAsFactors = FALSE
)

pca2_df <- merge(pca2_df, fam2, by = "sample.ID", all.x = TRUE)

#  PLOT (colored by superpop) 
sp <- factor(pca2_df$Super.Population)
plot(pca2_df$PC1, pca2_df$PC2,
     col = sp, pch = 19,
     xlab = "PC1", ylab = "PC2",
     main = paste0("PCA of CYP2D6 Locus Region"))
legend("topright", legend = levels(sp),
       col = seq_along(levels(sp)), pch = 19, cex = 0.8)
```


# 1b. PCA on Subpopulations (AFR and EAS) 
--> identify most separated subpopulations
```{r}
# SUBSET TO AFR + EAS INDIVIDUALS
pops_keep <- c("AFR", "EAS")
sub_ids <- fam2$sample.ID[fam2$Super.Population %in% pops_keep]
ind_rows <- which(fam$sample.ID %in% sub_ids)

# PCA RECOMPUTED ON ONLY AFR+EAS INDIVIDUALS 
# reuse the SAME SNP set ind.locus2 from Script 2 (important)
X_sub <- as.matrix(G[ind_rows, ind.locus2])

keep <- apply(X_sub, 2, var) > 0
X_sub <- X_sub[, keep, drop = FALSE]
X_sub <- scale(X_sub)

pca_sub <- prcomp(X_sub, center = FALSE, scale. = FALSE)

# dataframe for individuals in AFR+EAS
pca_sub_df <- data.frame(
  sample.ID = fam$sample.ID[ind_rows],
  PC1 = pca_sub$x[,1],
  PC2 = pca_sub$x[,2],
  stringsAsFactors = FALSE
)

pca_sub_df <- merge(pca_sub_df, fam2, by = "sample.ID")

# POPULATION CENTROIDS (GBR/FIN/etc — but here only AFR+EAS pops exist)
centroids_pop <- aggregate(cbind(PC1, PC2) ~ Population,
                           data = pca_sub_df,
                           mean)

# add superpopulation label for each Population
centroids_pop <- merge(centroids_pop,
                       unique(fam2[, c("Population", "Super.Population")]),
                       by = "Population")

# PLOT CENTROIDS COLORED BY SUPERPOP
cols <- ifelse(centroids_pop$Super.Population == "AFR", "black", "red")

plot(centroids_pop$PC1, centroids_pop$PC2,
     col = cols, pch = 19, cex = 2,
     xlab = "PC1", ylab = "PC2",
     main = "Population-Level PCA (AFR vs EAS)")

text(centroids_pop$PC1, centroids_pop$PC2,
     labels = centroids_pop$Population,
     pos = 3, cex = 0.9)

legend("topright", legend = c("AFR", "EAS"),
       col = c("black", "red"), pch = 19)

# MOST SEPARATED AFR vs EAS POPULATION PAIR (centroid distance) 
afr_cent <- subset(centroids_pop, Super.Population == "AFR")
eas_cent <- subset(centroids_pop, Super.Population == "EAS")

dist_matrix <- matrix(NA, nrow = nrow(afr_cent), ncol = nrow(eas_cent))
rownames(dist_matrix) <- afr_cent$Population
colnames(dist_matrix) <- eas_cent$Population

for (i in 1:nrow(afr_cent)) {
  for (j in 1:nrow(eas_cent)) {
    dist_matrix[i, j] <- sqrt(
      (afr_cent$PC1[i] - eas_cent$PC1[j])^2 +
        (afr_cent$PC2[i] - eas_cent$PC2[j])^2
    )
  }
}

max_idx <- which(dist_matrix == max(dist_matrix), arr.ind = TRUE)
afr_pop <- rownames(dist_matrix)[max_idx[1]]
eas_pop <- colnames(dist_matrix)[max_idx[2]]

cat("Most separated AFR–EAS pair:", afr_pop, "vs", eas_pop, "\n")
cat("Distance =", max(dist_matrix), "\n")


```



2. Allele Frequency Heatmaps
```{r}
ind.cyp <- which(map$chromosome == 22 &
                 map$physical.pos >= start_bp &
                 map$physical.pos <= end_bp)
length(ind.cyp)

maf <- bigsnpr::snp_MAF(G, ind.col = ind.cyp)
# keep informative SNPs
ind.cyp2 <- ind.cyp[maf >= 0.05 & maf <= 0.95]
cat("SNPs retained:", length(ind.cyp2), "\n")
```

```{r}
# extract genotype submatrix
X <- as.matrix(G[, ind.cyp2])

# add population labels
pop_labels <- fam2$Super.Population[
  match(fam$sample.ID, fam2$sample.ID)
]
```

# 2a. Allele Frequency Heatmap (Superpopulations)
```{r}
# compute allele frequency per superpopulation
pop_levels <- unique(pop_labels)

af_matrix <- matrix(NA,
                    nrow = length(pop_levels),
                    ncol = ncol(X))

rownames(af_matrix) <- pop_levels
colnames(af_matrix) <- paste0("SNP_", 1:ncol(X))

for (p in pop_levels) {
  inds <- which(pop_labels == p)
  af_matrix[p, ] <- colMeans(X[inds, ], na.rm = TRUE) / 2
}
```
Heatmap (all superpopulations)
```{r}
#install.packages("pheatmap")
library(pheatmap)

pheatmap(af_matrix,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         color = colorRampPalette(c("navy", "white", "firebrick"))(100),
         main = "CYP2D6 Allele Frequency Heatmap\nSuperpopulations")
```
Most Differentiated = AFR &	EAS
```{r}
# pairwise population differences
pop_diff <- matrix(0, nrow = nrow(af_matrix), ncol = nrow(af_matrix))
rownames(pop_diff) <- rownames(af_matrix)
colnames(pop_diff) <- rownames(af_matrix)

for(i in 1:nrow(af_matrix)){
  for(j in 1:nrow(af_matrix)){
    pop_diff[i,j] <- mean(abs(af_matrix[i,] - af_matrix[j,]), na.rm = TRUE)
  }
}

pop_diff

pop_pairs <- as.data.frame(as.table(pop_diff))
pop_pairs <- pop_pairs[pop_pairs$Var1 != pop_pairs$Var2, ]
pop_pairs <- pop_pairs[!duplicated(t(apply(pop_pairs[,1:2], 1, sort))), ]
# sort by difference
pop_pairs <- pop_pairs[order(-pop_pairs$Freq), ]
pop_pairs[1, ]
```
Heatmap (AFR & EAS only)
```{r}
library(pheatmap)

# Subset matrix to AFR and EAS
af_matrix_subset <- af_matrix[c("AFR", "EAS"), ]

# Plot heatmap
pheatmap(af_matrix_subset,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         color = colorRampPalette(c("navy", "white", "firebrick"))(100),
         main = "CYP2D6 Allele Frequency Heatmap\nAFR vs EAS")
```

# 2b. Allele Frequency Heatmap (Subpopulations of AFR & EAS)
Extract AFR + EAS individuals
```{r}
# Subpopulation labels
pop_sub <- fam2$Population[
  match(fam$sample.ID, fam2$sample.ID)
]

super_pop <- fam2$Super.Population[
  match(fam$sample.ID, fam2$sample.ID)
]

# keep AFR + EAS individuals
keep_inds_sub <- which(super_pop %in% c("AFR", "EAS"))

X_sub <- as.matrix(G[keep_inds_sub, ind.cyp2])

pop_sub_keep <- pop_sub[keep_inds_sub]
```
Compute allele frequency matrix
```{r}
subpop_levels <- unique(pop_sub_keep)

af_matrix_sub <- matrix(NA,
                        nrow = length(subpop_levels),
                        ncol = ncol(X_sub))

rownames(af_matrix_sub) <- subpop_levels
colnames(af_matrix_sub) <- paste0("SNP_", 1:ncol(X_sub))

for (p in subpop_levels) {
  inds <- which(pop_sub_keep == p)
  af_matrix_sub[p, ] <- colMeans(X_sub[inds, ], na.rm = TRUE) / 2
}

# remove low-variance SNPs
var_keep_sub <- apply(af_matrix_sub, 2, var) > 0.001
af_matrix_sub <- af_matrix_sub[, var_keep_sub]
```
Heatmap (all subpopulations)
```{r}
library(pheatmap)

pheatmap(af_matrix_sub,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         color = colorRampPalette(c("navy", "white", "firebrick"))(100),
         main = "CYP2D6 Allele Frequency\nAFR vs EAS Subpopulations")
```
Heatmap (ESN & CDX only)
```{r}
# Subset allele frequency matrix to ESN and CDX only
af_matrix_focus <- af_matrix_sub[rownames(af_matrix_sub) %in% c("ESN", "CDX"), ]

# Plot heatmap
pheatmap(af_matrix_focus,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         color = colorRampPalette(c("navy", "white", "firebrick"))(100),
         main = "CYP2D6 Allele Frequency: ESN vs CDX")
```

Most Differentiated = ESN & CDX
```{r}
# pairwise differences for subpopulations
pop_diff_sub <- matrix(0,
                       nrow = nrow(af_matrix_sub),
                       ncol = nrow(af_matrix_sub))

rownames(pop_diff_sub) <- rownames(af_matrix_sub)
colnames(pop_diff_sub) <- rownames(af_matrix_sub)

for(i in 1:nrow(af_matrix_sub)){
  for(j in 1:nrow(af_matrix_sub)){
    pop_diff_sub[i,j] <- mean(abs(af_matrix_sub[i,] - af_matrix_sub[j,]), na.rm = TRUE)
  }
}

pop_pairs_sub <- as.data.frame(as.table(pop_diff_sub))

# remove self comparisons
pop_pairs_sub <- pop_pairs_sub[pop_pairs_sub$Var1 != pop_pairs_sub$Var2, ]

# remove duplicate pairs (A-B vs B-A)
pop_pairs_sub <- pop_pairs_sub[!duplicated(t(apply(pop_pairs_sub[,1:2],1,sort))), ]

# sort by largest difference
pop_pairs_sub <- pop_pairs_sub[order(-pop_pairs_sub$Freq), ]
pop_pairs_sub[1, ]
```



3. Bootstrap Resampling for Cluster Stability (ARI)
Each bootstrap: Resample individuals → Recompute PCA → Recluster → Compare with truth
Then we compute: CI95% = [Q25%, Q75%]
```{r}
library(mclust)
set.seed(123)

B <- 1000   # use 10000 for final submission
n <- length(ind_rows)

# ORIGINAL PCA + CLUSTERING
X_orig <- as.matrix(G[ind_rows, ind.locus2])
keep <- apply(X_orig, 2, var) > 0
X_orig <- X_orig[, keep, drop = FALSE]
X_orig <- scale(X_orig)

pca_orig <- prcomp(X_orig, center = FALSE, scale. = FALSE)

# Use first 2 PCs for clustering
pc_orig <- pca_orig$x[, 1:2]

# k-means clustering (k=2)
km_orig <- kmeans(pc_orig, centers = 2, nstart = 20)
orig_clusters <- km_orig$cluster

# True labels (AFR vs EAS)
true_labels <- fam2$Super.Population[match(fam$sample.ID[ind_rows], fam2$sample.ID)]

# Convert AFR/EAS to numeric labels
true_labels_num <- as.numeric(factor(true_labels))

# ARI between original clustering and true labels
ari_original <- adjustedRandIndex(orig_clusters, true_labels_num)
cat("Original ARI:", ari_original, "\n")
```
Bootstrap
```{r}
ari_vals <- numeric(B)
for (b in 1:B) {
  # resample individuals with replacement
  boot_idx <- sample(1:n, n, replace = TRUE)
  
  X_boot <- X_orig[boot_idx, ]
  
  # recompute PCA
  pca_boot <- prcomp(X_boot, center = FALSE, scale. = FALSE)
  pc_boot <- pca_boot$x[, 1:2]
  
  # recluster
  km_boot <- kmeans(pc_boot, centers = 2, nstart = 10)
  boot_clusters <- km_boot$cluster
  
  # true labels for bootstrap sample
  boot_true <- true_labels_num[boot_idx]
  
  # compute ARI
  ari_vals[b] <- adjustedRandIndex(boot_clusters, boot_true)
}
```
Summary Statistics
```{r}
mean_ari <- mean(ari_vals)
ci <- quantile(ari_vals, c(0.025, 0.975))

cat("Bootstrap Mean ARI:", mean_ari, "\n")
cat("95% CI:", ci[1], "-", ci[2], "\n")

# Plot Distribution
hist(ari_vals, breaks = 40,
     main = "Bootstrap Distribution of ARI (AFR vs EAS)",
     xlab = "Adjusted Rand Index",
     col = "lightblue", border = "white")

abline(v = ci, col = "red", lwd = 2)
abline(v = mean_ari, col = "darkblue", lwd = 2)
```




BONUS: UMAP
```{r}
library(uwot)
set.seed(123)

# Use same scaled genotype matrix from PCA
umap_res <- umap(
  X,s
  n_neighbors = 15,
  min_dist = 0.1,
  metric = "euclidean"
)

umap_df <- data.frame(
  sample.ID = fam$sample.ID,
  UMAP1 = umap_res[,1],
  UMAP2 = umap_res[,2],
  stringsAsFactors = FALSE
)

umap_df <- merge(umap_df, fam2, by = "sample.ID")
```

```{r}
sp <- factor(umap_df$Super.Population)

plot(umap_df$UMAP1, umap_df$UMAP2,
     col = sp,
     pch = 19,
     xlab = "UMAP1",
     ylab = "UMAP2",
     main = "UMAP of CYP2D6 Region (Superpopulations)")

legend("topright",
       legend = levels(sp),
       col = seq_along(levels(sp)),
       pch = 19)
```




4. CYP2D6 Copy Number Variation (CNV) Analysis
--> see if CYP2D6 functional difference matches the genetic separation.
```{bash}
# Download 1000 Genomes SV dataset (which includes CNV info)
curl -O ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase3/integrated_sv_map/ALL.wgs.mergedSV.v8.20130502.svs.genotypes.vcf.gz
curl -O ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase3/integrated_sv_map/ALL.wgs.mergedSV.v8.20130502.svs.genotypes.vcf.gz.tbi

# Extract CYP2D6 region
bcftools view \
-r 22:41600000-43300000 \
ALL.wgs.mergedSV.v8.20130502.svs.genotypes.vcf.gz \
-Oz -o CYP2D6_SV2.vcf.gz

# index output
bcftools index CYP2D6_SV2.vcf.gz
```


```
library(vcfR)
vcf <- read.vcfR("CYP2D6_SV2.vcf.gz")
head(vcf@fix)
```

```
var_rows <- which(apply(gt, 1, function(x) length(unique(x)) > 1))
gt_var <- gt[var_rows, ]

cnv_numeric <- apply(gt_var, 2, function(x) {
  mean(ifelse(x == "0|0", 2,
       ifelse(x %in% c("0|1","1|0"), 3,
       ifelse(x == "1|1", 4, NA))), na.rm = TRUE)
})

cnv_df <- data.frame(
  sample.ID = colnames(gt),
  copy_number = cnv_numeric
)

cnv_df <- merge(cnv_df, fam2, by = "sample.ID")

cnv_focus <- cnv_df %>%
  dplyr::filter(Population %in% c("ESN", "CDX"))
```


cnv_summary <- cnv_df %>%
  filter(Super.Population %in% c("AFR","EAS")) %>%
  group_by(Population, Super.Population) %>%
  summarise(
    mean_CN = mean(copy_number),
    sd_CN = sd(copy_number),
    .groups="drop"
  )

ggplot(cnv_summary,
       aes(x = Population, y = mean_CN, color = Super.Population)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = mean_CN - sd_CN,
                    ymax = mean_CN + sd_CN),
                width = 0.2) +
  theme_minimal() +
  labs(
    title = "Mean CYP2D6 Copy Number by Population",
    y = "Mean Copy Number"
  )

# Generate synthetic datasets

library(data.table)
set.seed(1)

# Study-level constants
N      <- 150
N_male <- 75
N_fem  <- 75
SEX    <- c(rep(1, N_male), rep(0, N_fem))

WT <- numeric(N)
WT[SEX==1] <- pmax(rnorm(sum(SEX==1), mean=71, sd=9.9), 40)
WT[SEX==0] <- pmax(rnorm(sum(SEX==0), mean=50, sd=5.1), 35)

p_meal   <- 0.68
EMPTY    <- rbinom(N, size=1, prob=1 - p_meal)
DOSE_GKG <- runif(N, min=0.5, max=0.7)
D_g      <- DOSE_GKG * WT    # grams ethanol
D_mg     <- D_g * 1000        # mg ethanol — KEY FIX

# Yang 2011 parameters
b1      <- 4.283
b2      <- 2.100
b3      <- -0.880
b_wt    <- 0.015
b_empty <- 0.194
b_sex   <- -0.154

Omega <- matrix(c(
  0.209, 0.000, 0.173,
  0.000, 0.409, 0.000,
  0.173, 0.000, 0.192
), nrow=3, byrow=TRUE)

L   <- chol(Omega)
Z   <- matrix(rnorm(N*3), ncol=3)
ETA <- Z %*% L
eta1 <- ETA[,1]
eta2 <- ETA[,2]
eta3 <- ETA[,3]

# Cl in dL/hr — with D in mg -> cp in mg/100mL
Cl  <- exp(b1 + b_wt * WT + eta1)
ka  <- exp(b2 + b_empty * EMPTY + eta2)
kel <- exp(b3 + b_sex * SEX + eta3)

# Sampling times
t_early <- seq(0.17, 1.0, by=0.17)
t_late  <- seq(1.5, 6.0, by=0.5)   # extended to 6hr to match Nigerian
times   <- c(0, t_early, t_late)

# Concentration function
#    D in mg, Cl in dL/hr -> cp in mg/100mL
Ct_fun <- function(t, D, Cl, ka, kel){
  if (abs(ka - kel) < 1e-6) return(rep(0, length(t)))
  D * ka * kel / (Cl * (ka - kel)) * (exp(-kel*t) - exp(-ka*t))
}

# Residual error on mg/100mL scale
sigma_mg <- sqrt(11.46)   # ~3.38 mg/100mL (Yang et al.)
LLOQ     <- 0.2           # mg/100mL

# Build dataset
out <- rbindlist(lapply(1:N, function(i){
  conc <- sapply(times, Ct_fun, D=D_mg[i], Cl=Cl[i], ka=ka[i], kel=kel[i])
  dv   <- pmax(conc + rnorm(length(times), 0, sigma_mg), 0)
  dv[dv < LLOQ] <- 0
  data.table(
    ID       = i,
    TIME     = times,
    DV       = dv,         # mg/100mL
    SEX      = SEX[i],
    WT       = WT[i],
    EMPTY    = EMPTY[i],
    DOSE_GKG = DOSE_GKG[i]
  )
}))

out[, `:=`(AMT=0, EVID=0, CMT=1)]
out[TIME==0, `:=`(AMT = D_mg[ID], EVID=1, DV=NA_real_)]  # AMT in mg

# verify
cat("AMT range (mg):       ", round(range(out[EVID==1, AMT])), "\n")
cat("DV range (mg/100mL):  ", round(range(out[EVID==0 & DV>0, DV]), 1), "\n")

fwrite(out, "synthetic_chinese_popPK.csv")
out[]
```

library(data.table)
set.seed(3)

# Study-level constants
N      <- 150
N_male <- 75
SEX    <- c(rep(1, N_male), rep(0, N - N_male))  # 1=male, 0=female
WT     <- pmax(rnorm(N, mean=55, sd=3), 40)

p_meal   <- 0.68
EMPTY    <- rbinom(N, size=1, prob = 1 - p_meal)  # 1=empty stomach
DOSE_GKG <- runif(N, min=0.5, max=0.7)
D_g      <- DOSE_GKG * WT     # grams ethanol
D_mg     <- D_g * 1000         # mg ethanol

#  Individual blood volume via Widmark r factor
r_widmark    <- ifelse(SEX == 1, 0.68, 0.55)  # L/kg
blood_vol_dL <- r_widmark * WT * 10            # dL (100 mL units)

# Ct model
#    Units: D in mg, Cl in dL/hr -> conc in mg/100mL

Ct_fun <- function(t, D, Cl, ka, kel){
  if (abs(ka - kel) < 1e-6) return(rep(0, length(t)))
  D * ka * kel / (Cl * (ka - kel)) * (exp(-kel*t) - exp(-ka*t))
}

# Helper functions
rtnorm_pos <- function(n, mean, sd, min = 1e-6){
  x <- rnorm(n, mean, sd)
  while(any(x <= min)){
    x[x <= min] <- rnorm(sum(x <= min), mean, sd)
  }
  x
}

solve_ka <- function(kel, tmax){
  f <- function(ka) log(ka / kel) / (ka - kel) - tmax
  uniroot(f, lower = kel + 1e-4, upper = 200)$root
}

solve_cl_from_cmax <- function(D, ka, kel, tmax, cmax){
  num <- D * ka * kel / (ka - kel) * (exp(-kel * tmax) - exp(-ka * tmax))
  Cl  <- num / cmax
  if (!is.finite(Cl) || Cl <= 0) NA_real_ else Cl
}

# PK targets — adjusted for fast Nigerian clearance
# Lower Cmax + faster kel -> higher Cl -> faster clearance
cmax_m_mean <- 70;   cmax_m_sd <- 4     # mg/100mL (reduced from 90)
cmax_f_mean <- 75;   cmax_f_sd <- 5     # mg/100mL (reduced from 96)

tmax_m_mean <- 0.60; tmax_m_sd <- 0.12  # hrs (faster than SC)
tmax_f_mean <- 0.45; tmax_f_sd <- 0.10

kel_m_mean  <- 0.35; kel_m_sd  <- 0.04  # /hr (half-life ~2 hr)
kel_f_mean  <- 0.40; kel_f_sd  <- 0.04  # /hr (half-life ~1.7 hr)

#  Generate individual PK parameters
kel  <- numeric(N)
tmax <- numeric(N)
cmax <- numeric(N)

kel[SEX==1]  <- rtnorm_pos(sum(SEX==1), kel_m_mean, kel_m_sd)
kel[SEX==0]  <- rtnorm_pos(sum(SEX==0), kel_f_mean, kel_f_sd)
tmax[SEX==1] <- rtnorm_pos(sum(SEX==1), tmax_m_mean, tmax_m_sd, min=0.05)
tmax[SEX==0] <- rtnorm_pos(sum(SEX==0), tmax_f_mean, tmax_f_sd, min=0.05)
cmax[SEX==1] <- rtnorm_pos(sum(SEX==1), cmax_m_mean, cmax_m_sd)
cmax[SEX==0] <- rtnorm_pos(sum(SEX==0), cmax_f_mean, cmax_f_sd)

ka <- mapply(solve_ka, kel=kel, tmax=tmax)
Cl <- mapply(function(D, ka, kel, tmax, cmax){
  solve_cl_from_cmax(D, ka, kel, tmax, cmax)
}, D=D_mg, ka=ka, kel=kel, tmax=tmax, cmax=cmax)

# sanity check
cat("Cl range (dL/hr):          ", round(range(Cl), 1), "\n")
cat("kel range (/hr):           ", round(range(kel), 3), "\n")
cat("Implied half-life range (hr):", round(range(log(2)/kel), 2), "\n")
cat("ka range (/hr):            ", round(range(ka), 2), "\n")

stopifnot(all(is.finite(Cl)), all(Cl > 0))

# Sampling times
t_early <- seq(0.17, 1.0, by=0.17)
t_late  <- seq(1.5, 6.0, by=0.5)
times   <- c(0, t_early, t_late)

# Residual error (mg/100mL scale)
sigma_mg <- sqrt(11.46)  # ~3.38 mg/100mL (Yang et al.)
cv       <- 0.05         # 5% proportional
LLOQ     <- 0.2          # mg/100mL

# Build dataset
out <- rbindlist(lapply(1:N, function(i){
  conc <- sapply(times, Ct_fun, D=D_mg[i], Cl=Cl[i], ka=ka[i], kel=kel[i])
  dv   <- conc * (1 + rnorm(length(times), 0, cv)) + rnorm(length(times), 0, sigma_mg)
  dv   <- pmax(dv, 0)
  dv[dv < LLOQ] <- 0
  data.table(
    ID           = i,
    TIME         = times,
    DV           = dv,
    SEX          = SEX[i],
    WT           = WT[i],
    EMPTY        = EMPTY[i],
    DOSE_GKG     = DOSE_GKG[i],
    BLOOD_VOL_dL = blood_vol_dL[i]
  )
}))

out[, `:=`(AMT=0, EVID=0, CMT=1)]
out[TIME==0, `:=`(AMT = D_mg[ID], EVID=1, DV=NA_real_)]

# verify output looks right
cat("\nAMT range (mg):        ", round(range(out[EVID==1, AMT])), "\n")
cat("DV range (mg/100mL):   ", round(range(out[EVID==0 & DV>0, DV]), 1), "\n")

fwrite(out, "synthetic_nigerian_popPK.csv")
out[]
```


```{r}
library(data.table)
library(ggplot2)

# Shared model parameters (Yang 2011)
b1 <- 4.283
b2 <- 2.100
b3 <- -0.880
b_wt    <- 0.015
b_empty <- 0.194
b_sex   <- -0.154

Omega <- matrix(c(
  0.209, 0.000, 0.173,
  0.000, 0.409, 0.000,
  0.173, 0.000, 0.192
), nrow=3, byrow=TRUE)

L <- chol(Omega)

Ct_fun <- function(t, D, Cl, ka, kel){
  if (abs(ka - kel) < 1e-9) return(rep(0, length(t)))
  D * ka * kel / (Cl * (ka - kel)) * (exp(-kel*t) - exp(-ka*t))
}

# smooth curve time grid
times_smooth <- seq(0, 24, by = 0.02)

# observed sampling time grid (like your dataset)
t_early <- seq(0.17, 5.0, by=0.17)
t_late  <- seq(5.0, 10.0 , by=0.5)
times_obs <- c(0.17, t_early, t_late) # exclude 0 dose time for plotting

# Simulator: now supports CL multipliers + kel multipliers
simulate_conc_pop <- function(
  N,
  seed,
  wt_fun,
  dose_gkg_fun,
  empty_fun,
  cl_multiplier_fun,     
  kel_multiplier_fun
){
  set.seed(seed)

  N_male <- floor(N/2)
  N_fem  <- N - N_male
  SEX <- c(rep(1, N_male), rep(0, N_fem))

  WT <- wt_fun(SEX)
  DOSE_GKG <- dose_gkg_fun(N)
  D_g <- DOSE_GKG * WT
  EMPTY <- empty_fun(N)

  Z <- matrix(rnorm(N*3), ncol=3)
  ETA <- Z %*% L
  eta1 <- ETA[,1]; eta2 <- ETA[,2]; eta3 <- ETA[,3]

  Cl_base  <- exp(b1 + b_wt * WT + eta1)
  ka       <- exp(b2 + b_empty * EMPTY + eta2)
  kel_base <- exp(b3 + b_sex * SEX + eta3)

  # apply population-level multipliers
  Cl  <- Cl_base  * cl_multiplier_fun(SEX)
  kel <- kel_base * kel_multiplier_fun(SEX)

  conc_mat <- matrix(NA_real_, nrow = N, ncol = length(times_smooth))
  for(i in 1:N){
    conc_mat[i,] <- Ct_fun(times_smooth, D_g[i], Cl[i], ka[i], kel[i])
  }

  list(conc = conc_mat, SEX=SEX, WT=WT, DOSE_GKG=DOSE_GKG, EMPTY=EMPTY,
       D_g=D_g, Cl=Cl, ka=ka, kel=kel)
}

# Population definitions
wt_yang <- function(SEX){
  N <- length(SEX)
  WT <- numeric(N)
  WT[SEX==1] <- pmax(rnorm(sum(SEX==1), mean=71, sd=9.9), 40)
  WT[SEX==0] <- pmax(rnorm(sum(SEX==0), mean=50, sd=5.1), 35)
  WT
}

wt_nig <- function(SEX){
  N <- length(SEX)
  pmax(rnorm(N, mean=55, sd=3), 40)
}

dose_like_original <- function(N) runif(N, min=0.5, max=0.7)

empty_like_original <- function(N){
  p_meal <- 0.68
  rbinom(N, 1, prob = 1 - p_meal)
}

# Multipliers
# Southern Chinese: baseline
cl_mult_yang  <- function(SEX) rep(1, length(SEX))
kel_mult_yang <- function(SEX) rep(1, length(SEX))

# Nigerian: increase CL to reduce exposure WITHOUT spiky peak
nigerian_cl_multiplier <- 2.3  # <-- tune: try 1.2 to 1.8
cl_mult_nig <- function(SEX) rep(nigerian_cl_multiplier, length(SEX))

# keep kel ~ same as baseline (optional female ratio)
nigerian_kel_multiplier <- 1.0
female_extra_multiplier <- 0.026/0.021
kel_mult_nig <- function(SEX){
  base <- rep(nigerian_kel_multiplier, length(SEX))
  base[SEX==0] <- base[SEX==0] * female_extra_multiplier
  base
}

# Simulate populations
yang <- simulate_conc_pop(
  N = 2000, seed = 1,
  wt_fun = wt_yang,
  dose_gkg_fun = dose_like_original,
  empty_fun = empty_like_original,
  cl_multiplier_fun = cl_mult_yang,
  kel_multiplier_fun = kel_mult_yang
)

nig <- simulate_conc_pop(
  N = 2000, seed = 2,
  wt_fun = wt_nig,
  dose_gkg_fun = dose_like_original,
  empty_fun = empty_like_original,
  cl_multiplier_fun = cl_mult_nig,
  kel_multiplier_fun = kel_mult_nig
)

# Summarize smooth mean + 5-95% band

summarize_profile <- function(conc_mat, label){
  data.table(
    time = times_smooth,
    mean = colMeans(conc_mat),
    p05  = apply(conc_mat, 2, quantile, probs=0.05),
    p95  = apply(conc_mat, 2, quantile, probs=0.95),
    pop  = label
  )
}

prof <- rbind(
  summarize_profile(yang$conc, "Southern Chinese"),
  summarize_profile(nig$conc,  "Nigerian")
)

# Keep observed points (unchanged logic)

make_obs5 <- function(sim, pop_label, seed=999, meas_sd=0.008){
  set.seed(seed)
  id <- sample(seq_along(sim$D_g), 1)

  pred <- Ct_fun(times_obs, sim$D_g[id], sim$Cl[id], sim$ka[id], sim$kel[id])

  idx <- sample(seq_along(times_obs), 5)
  dt <- data.table(
    time = times_obs[idx],
    conc = pmax(pred[idx] + rnorm(5, 0, meas_sd), 0),
    pop  = pop_label
  )
  dt[order(time)]
}

obs_ch5 <- make_obs5(yang, "Southern Chinese", seed=1001, meas_sd=0.008)
obs_ng5 <- make_obs5(nig,  "Nigerian",        seed=1002, meas_sd=0.008)
obs5 <- rbind(obs_ch5, obs_ng5)

# Plot: smooth line + band + points
p <- ggplot(prof, aes(x=time, y=mean, color=pop, fill=pop)) +
  geom_ribbon(aes(ymin=p05, ymax=p95), alpha=0.15, color=NA) +
  geom_line(linewidth=1.1) +
  geom_point(
    data = obs5,
    aes(x=time, y=conc, shape=pop),
    inherit.aes = FALSE,
    color = "black",
    size = 3
  ) +
  labs(x = "Time (h)", y = "BAC (mg/100 mL)") +
  coord_cartesian(xlim=c(0, 12)) +
  theme_minimal(base_size = 14) +
  theme(legend.title = element_blank())

p
```



library(data.table)
library(nlme)
library(ggplot2)

# Closed-form 1cpt oral model
cp_oral_1cpt <- function(time, dose, lcl, lka, lkel) {
  cl  <- exp(lcl)
  ka  <- exp(lka)
  kel <- exp(lkel)
  denom <- (ka - kel)

  pred <- ifelse(abs(denom) < 1e-8,
                 0,
                 dose * ka * kel / (cl * denom) * (exp(-kel*time) - exp(-ka*time)))
  # prevent log(0)
  pmax(pred, 1e-12)
}

# log-pred wrapper (proportional error model on original scale)
lpred_wrapper <- function(TIME, DOSE_G,
                          lcl_0, lcl_wt,
                          lka_0, lka_empty,
                          lkel_0, lkel_sex,
                          WTc, EMPTY, SEX) {
  pred <- cp_oral_1cpt(
    TIME, DOSE_G,
    lcl  = lcl_0 + lcl_wt * WTc,
    lka  = lka_0 + lka_empty * EMPTY,
    lkel = lkel_0 + lkel_sex * SEX
  )
  log(pred)
}

ctrl <- nlmeControl(
  pnlsTol = 1e-6, msTol = 1e-6,
  maxIter = 300, pnlsMaxIter = 80, msMaxIter = 300,
  returnObject = TRUE
)
# Load + prep
load_pop <- function(path, label){
  dt <- fread(path)
  dt <- dt[EVID == 0 & !is.na(DV)]
  dt[, POP := label]
  dt[, ID := as.factor(ID)]
  dt[, SEX := as.numeric(SEX)]
  dt[, EMPTY := as.numeric(EMPTY)]
  dt[, WT := as.numeric(WT)]
  dt[, DOSE_G := DOSE_GKG * WT]
  dt
}

# Fit one population robustly
fit_one_pop <- function(dt, label, blq = c("drop0","replace0")) {
  blq <- match.arg(blq)
  cat("\n====================\n")
  cat("Fitting population:", label, "\n")
  cat("====================\n")

  # Center WT within dataset
  muWT <- mean(dt$WT, na.rm=TRUE)
  dt[, WTc := WT - muWT]

  # BLQ handling
  prop0 <- mean(dt$DV == 0, na.rm=TRUE)
  cat("Proportion DV==0:", round(prop0, 4), "\n")

  dt2 <- copy(dt)
  if(blq == "drop0"){
    dt2 <- dt2[DV > 0]
  } else {
    # replace zeros with half of small positive reference (5th percentile of positives)
    ref <- quantile(dt2$DV[dt2$DV > 0], probs=0.05, na.rm=TRUE)
    dt2[DV == 0, DV := as.numeric(ref)/2]
  }

  # log DV
  dt2[, LDV := log(DV)]

  # Starting values:
  # If WT is centered, intercept shifts by b_wt*mean(WT)
  start <- c(
    lcl_0     = 4.283 + 0.015 * muWT,
    lcl_wt    = 0.015,
    lka_0     = 2.100,
    lka_empty = 0.194,
    lkel_0    = -0.880,
    lkel_sex  = -0.154
  )

  fixed_names <- names(start)

  cat("Step 0: gnls fixed-only...\n")
  fit0 <- gnls(
    LDV ~ lpred_wrapper(TIME, DOSE_G,
                        lcl_0, lcl_wt,
                        lka_0, lka_empty,
                        lkel_0, lkel_sex,
                        WTc, EMPTY, SEX),
    data = dt2,
    start = start,
    control = ctrl
  )
  cat("gnls OK. logLik:", logLik(fit0), "\n")

  # ---- Step 1: nlme random CL only ----
  cat("Step 1: nlme random = pdDiag(CL)\n")
  st1 <- coef(fit0)[fixed_names]

  fit1 <- nlme(
    LDV ~ lpred_wrapper(TIME, DOSE_G,
                        lcl_0, lcl_wt,
                        lka_0, lka_empty,
                        lkel_0, lkel_sex,
                        WTc, EMPTY, SEX),
    data = dt2,
    fixed  = lcl_0 + lcl_wt + lka_0 + lka_empty + lkel_0 + lkel_sex ~ 1,
    random = pdDiag(lcl_0 ~ 1),
    groups = ~ ID,
    start  = list(fixed = st1),
    control = ctrl,
    na.action = na.omit
  )
  cat("Step 1 VarCorr:\n"); print(VarCorr(fit1))

  cat("Step 2: try random = pdDiag(CL, KEL)\n")
  fit2 <- try(update(fit1, random = pdDiag(lcl_0 + lkel_0 ~ 1)), silent=TRUE)
  if(inherits(fit2, "try-error")){
    cat("Step 2 failed; keeping Step 1.\n")
    return(list(fit = fit1, data = dt2, muWT = muWT, step = "1"))
  }
  cat("Step 2 VarCorr:\n"); print(VarCorr(fit2))

  cat("Step 3: try random = pdDiag(CL, KA, KEL)\n")
  fit3 <- try(update(fit2, random = pdDiag(lcl_0 + lka_0 + lkel_0 ~ 1)), silent=TRUE)
  if(inherits(fit3, "try-error")){
    cat("Step 3 failed; keeping Step 2.\n")
    return(list(fit = fit2, data = dt2, muWT = muWT, step = "2"))
  }

  # Guardrail for ka SD explosion
  vc <- VarCorr(fit3)
  ka_sd <- suppressWarnings(as.numeric(vc["lka_0","StdDev"]))
  if(is.na(ka_sd) || ka_sd > 5){
    cat("KA random SD looks degenerate (", ka_sd, "). Keeping Step 2.\n", sep="")
    return(list(fit = fit2, data = dt2, muWT = muWT, step = "2"))
  }

  cat("Step 3 VarCorr:\n"); print(VarCorr(fit3))
  return(list(fit = fit3, data = dt2, muWT = muWT, step = "3"))
}

# Run fits
sc <- load_pop("synthetic_chinese_popPK.csv", "SC")
ng <- load_pop("synthetic_nigerian_popPK.csv", "NG")

# If SC has tons of zeros, drop0 is usually best
res_sc <- fit_one_pop(sc, "SC", blq="drop0")
res_ng <- fit_one_pop(ng, "NG", blq="drop0")

fit_sc <- res_sc$fit
fit_ng <- res_ng$fit

cat("\n--- SC final step:", res_sc$step, "---\n")
print(summary(fit_sc))
cat("\n--- NG final step:", res_ng$step, "---\n")
print(summary(fit_ng))

# Compare fixed effects (and CIs)
get_fixed_ci <- function(fit, pop){
  fe <- fixef(fit)
  ci <- intervals(fit, which="fixed")$fixed
  data.table(
    POP = pop,
    param = names(fe),
    est = as.numeric(fe),
    lo  = ci[names(fe), "lower"],
    hi  = ci[names(fe), "upper"]
  )
}

dt_sc <- get_fixed_ci(fit_sc, "SC")
dt_ng <- get_fixed_ci(fit_ng, "NG")
dt_all <- rbind(dt_sc, dt_ng)

print(dt_all)

# CL fold-change NG/SC (intercept on centered WT)
cl_sc <- dt_sc[param=="lcl_0", est]
cl_ng <- dt_ng[param=="lcl_0", est]
cat("\nApprox CL fold-change NG/SC (exp(delta lcl_0)) =", exp(cl_ng - cl_sc), "\n")

# Plot key intercepts with CI
plot_dt <- dt_all[param %in% c("lcl_0","lka_0","lkel_0")]
plot_dt[, param := factor(param,
                          levels=c("lcl_0","lka_0","lkel_0"),
                          labels=c("log(CL) intercept","log(ka) intercept","log(kel) intercept"))]

ggplot(plot_dt, aes(x=param, y=est, color=POP)) +
  geom_point(position=position_dodge(width=0.4), size=3) +
  geom_errorbar(aes(ymin=lo, ymax=hi),
                width=0.2,
                position=position_dodge(width=0.4)) +
  labs(x="", y="Estimate (log-scale) with 95% CI",
       title="Separate nlme fits: SC vs NG (log-DV model)") +
  theme_minimal(base_size=13) +
  theme(axis.text.x=element_text(angle=20, hjust=1),
        legend.title=element_blank())



library(data.table)

sc_obs <- fread("synthetic_chinese_popPK.csv")
ng_obs <- fread("synthetic_nigerian_popPK.csv")
sc_obs[, POP := "Southern Chinese"]
ng_obs[, POP := "Nigerian"]
obs <- rbind(sc_obs, ng_obs, fill=TRUE)
obs <- obs[EVID == 0 & !is.na(DV)]

cmax_tbl <- obs[, .(Cmax = max(DV, na.rm=TRUE)), by=.(ID, POP)]

auc_tbl <- obs[, {
  dt  <- .SD[order(TIME)][TIME > 0]
  auc <- sum(diff(dt$TIME) * (head(dt$DV, -1) + tail(dt$DV, -1)) / 2)
  list(AUC = auc)
}, by=.(ID, POP)]

gof_sc <- as.data.table(fit_sc)
gof_ng <- as.data.table(fit_ng)
gof_sc[, POP := "Southern Chinese"]
gof_ng[, POP := "Nigerian"]
gof    <- rbind(gof_sc, gof_ng, fill=TRUE)
gof    <- gof[!is.na(DV) & !is.na(PRED) & PRED > 0]
cv_tbl <- gof[, .(
  CV_pct = round(sd(DV - PRED, na.rm=TRUE) / mean(PRED, na.rm=TRUE) * 100, 1)
), by=POP]

pk_tbl <- merge(cmax_tbl, auc_tbl, by=c("ID", "POP"))

summary_tbl <- pk_tbl[, .(
  Cmax = round(mean(Cmax), 1),
  AUC  = round(mean(AUC),  1)
), by=POP]

summary_tbl <- merge(summary_tbl, cv_tbl, by="POP")
setnames(summary_tbl, "CV_pct", "CV%")


print(summary_tbl)
```


